ODGI: understanding pangenome graphs

### 2021-11-01 manuscript/main.tex
the discussion is hard

\abstract{
\textbf{Motivation:}
Pangenome graphs provide a complete representation of the mutual alignment of collections of genomes.
These models offer the opportunity to study the entire genomic diversity of a population, including structurally complex regions.
Nevertheless, analyzing hundreds of gigabase-scale genomes using pangenome graphs is difficult as it is not well-supported by existing tools.
Hence, fast and versatile software is required to ask advanced questions to such data in an efficient way. \\
\textbf{Results:}
We wrote ODGI, a novel suite of tools that implements scalable algorithms and has an efficient in-memory representation of DNA variation graphs.
ODGI includes tools for detecting complex regions, extracting \textit{loci}, removing artifacts, exploratory analysis, manipulation, validation, and visualization.
Its fast parallel execution facilitates routine pangenomic tasks as well as pipelines that can quickly answer complex biological questions of gigabase-scale pangenome graphs. \\
\textbf{Availability:}
ODGI is published as free software under the MIT open source license.
Source code can be downloaded from \url{https://github.com/pangenome/odgi} and documentation is available at \url{https://odgi.readthedocs.io}.
ODGI can be installed via Bioconda \url{https://bioconda.github.io/recipes/odgi/README.html} or GNU Guix \url{https://github.com/ekg/guix-genomics/blob/master/odgi.scm}. \\
\textbf{Contact:} \href{egarris5@uthsc.edu}{egarris5@uthsc.edu} \\
\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics} online.
}

@@

\section{Discussion}

Pangenome graphs stand to become a ubiquitous tool in genomics~\citep{Eizenga_2020}.
With ODGI we implemented a state-of-the-art tool suite that can transform, analyse, simplify, validate, and visualise pangenome graphs at large scale.

Lifting over annotations and linearizing nested graph structures place the suite as the bridge between traditional linear reference genome analysis and pangenome graphs.
ODGI is a unique set of tools that enables scientists to explore and discover the underlying biology of pangenome graphs.
Already, the tools are the backbone of pipelines such as the Pangenome Graph Builder~\citep{pggb} (PGGB) or nf-core/pangenome~\citep{pangenome}. %\footnote{\url{https://github.com/nf-core/pangenome} (accessed Oct 2021)}.
Future work will add support for RNA and protein sequences and expand on metadata capabilities of large pangenome graphs.

ODGI provides common operation on standard pangenomic file formats, such as GFA.
ODGI aims to be a reference implementation for many pangenome related operations.
In the near future we aim to improve RDF support for annotation in the in-memory graph~\citep{Yokoyama2020}, for example, and allow for federated queries on the pangenome, and FAIR data sharing.

ODGI's visualization tools produce static, large-scale snapshots.
Recent interactive browsers are reference centric~\citep{Beyer2019, Yokoyama2019, Durant2021, Liang2021} or focusing on one dimension~\citep{Wick_2015, Gonnella2018}.
However, an interactive visualization solution that combines the 1D and 2D layout of a graph with annotation and read mapping information across different zoom levels is still missing.
ODGI's graph layout algorithms could become a backbone of such tools, as is already the case for \textit{gfaestus} \citep{gfaestus}%\footnote{\url{https://github.com/chfi/gfaestus}}, which currently relies on the \textit{odgi layout} output for its interactive 2D visualization.

@@

With the increased adoption of long read sequencing we expect pangenome tools to become increasingly common in biomedical research.
Particularly for targets that involve complex variation, such as cancer, plant genomics and metagenomics, ODGI should facilitate disentangling, describing and analysing a much larger set of variation than previously was possible with tools that depend on short reads and reference genomes.

@@

Data used to build Human pangenome graphs is available at \url{https://github.com/human-pangenomics/HPP_Year1_Data_Freeze_v1.0}.
TODO!

>>>>

\abstract{
\textbf{Motivation:}
Pangenome graphs provide a complete representation of the mutual alignment of collections of genomes.
These models offer the opportunity to study the entire genomic diversity of a population, including structurally complex regions.
Nevertheless, analyzing hundreds of gigabase-scale genomes using pangenome graphs is difficult as it is not well-supported by existing tools.
Hence, fast and versatile software is required to ask advanced questions to such data in an efficient way. \\
\textbf{Results:}
We wrote ODGI, a novel suite of tools that implements scalable algorithms and has an efficient in-memory representation of DNA variation graphs.
ODGI includes tools for detecting complex regions, extracting \textit{loci}, removing artifacts, exploratory analysis, manipulation, validation, and visualization.
Its fast parallel execution facilitates routine pangenomic tasks as well as pipelines that can quickly answer complex biological questions of gigabase-scale pangenome graphs. \\
\textbf{Availability:}
ODGI is published as free software under the MIT open source license.
Source code can be downloaded from \url{https://github.com/pangenome/odgi} and documentation is available at \url{https://odgi.readthedocs.io}.
ODGI can be installed via Bioconda \url{https://bioconda.github.io/recipes/odgi/README.html} or GNU Guix \url{https://github.com/ekg/guix-genomics/blob/master/odgi.scm}. \\
\textbf{Contact:} \href{egarris5@uthsc.edu}{egarris5@uthsc.edu} \\
%\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics} online.
}

@@

\section{Discussion}
% Short recap
% Confessions
% Future plans
Pangenome graphs stand to become a ubiquitous model in genomics~\citep{Eizenga_2020}.
With ODGI we implemented a state-of-the-art tool suite that enables users to explore and discover the underlying biology in pangenomes graphs, filling the gap that made pangenomic analyzes difficult.
Indeed, with ODGI's tools users can easily transform, analyse, simplify, validate, and visualise pangenome graphs at large scale.

ODGI's visualization tools produce static, large-scale snapshots.
Recent interactive browsers are reference centric~\citep{Beyer2019, Yokoyama2019, Durant2021, Liang2021} or focusing on 1D~\citep{Wick_2015, Gonnella2018}.
However, an interactive visualization solution that combines the 1D and 2D layout of a graph with annotation and read mapping information across different zoom levels is still missing.
Nevertheless, ODGI's graph layout algorithms could become a backbone of such tools, as is already the case for \textit{gfaestus}~\citep{gfaestus} %\footnote{\url{https://github.com/chfi/gfaestus}},
which currently relies on the \textit{odgi layout} output for its interactive 2D visualization.

Lifting over annotations and linearizing nested graph structures place the suite as the bridge between traditional linear reference genome analysis and pangenome graphs.

Already, the tools are the backbone of pipelines such as the Pangenome Graph Builder~\citep{pggb} (PGGB) or nf-core/pangenome~\citep{pangenome}. %\footnote{\url{https://github.com/nf-core/pangenome} (accessed Oct 2021)}.
Future work will add support for RNA and protein sequences and expand on metadata capabilities of large pangenome graphs.

ODGI aims to be a reference implementation for many pangenome related operations.
In the near future we aim to improve RDF support for annotation in the in-memory graph~\citep{Yokoyama2020}, for example, and allow for federated queries on the pangenome, and FAIR data sharing.

@@

With the increased adoption of long read sequencing we expect pangenome tools to become increasingly common in biomedical research.
Particularly for targets that involve complex variation, such as cancer, plant genomics and metagenomics, ODGI will facilitate disentangling,
describing and analysing a much larger set of variation than previously was possible with tools that depend on short reads and reference genomes.

@@

\FIXME{Data used to build Human pangenome graphs is available at \url{https://github.com/human-pangenomics/HPP_Year1_Data_Freeze_v1.0}.}
Code used to build images in this manuscript is available at \url{https://github.com/pangenome/odgi-paper}.
TODO!
